Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-08 paper/tidy-missing-data-paper.Rmd
touching up final section
addressing most of the notes

**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$. Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In **MNAR**, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.  This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR. Imputation is recommended in most cases of missingness dependence.

<!-- (NOTE: Are the above last two sentences OK?) -->
<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->

@@

<!-- Note: I'm not sure why I talk about this on the fly stuff here, and I wonder if this paragraph below could be removed? -->

The _nabular_ data structure provides three key benefits over creating a missingness matrix, $B$, on-the-fly. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these imputations.

@@

<!-- Note: I'm not sure where I should put these three parts below parts in the paper: -->

<!-- - Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.  SAS, SPSS, and STATA all implement. -->

<!-- - The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R. -->
<!-- - (in the same column you have mixed information, which breaks the tidy principle of one column having one type of value) -->

@@

<!-- note: Di can you check these last two paragraphs -->

@@

@@

The authors would like to thank Miles McBain, for his key contributions and discussions on the `naniar` package, in particular for helping implement `geom_miss_point`, and for his feedback on ideas, implementations, and names. We also thank Colin Fay for his contributions to the `naniar` package, in particular for his assistance with the `replace_with_na` functions. We also thank Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on missing data and package development, and for their assistance with creating elegant approaches that take advantage of the tidy syntax. We would also like to thank those who contributed pull requests and discussions on the `naniar` package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping `naniar` be more compliant with `ggplot2`. We would also like to thank Amelia McNamara for her assistance with proof-reading the paper.

>>>>

**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$. Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In **MNAR**, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.  This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR.

<!-- (NOTE: Is the last sentence OK?) -->
<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->

@@

<!-- Note: Do you think this paragraph needs to be here, about the use of _nabular_ data structure? -->

Although a binary missingness matrix, $B$, could be generated during analysis instead of using a _nabular_ data structure, there are key advantages to using _nabular_ data. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these imputations.

@@

Special missing values are not a new idea, and have been implemented in other statistical programming languages, SPSS, SAS, and STATA. These typically represent missing values as a full-stop, `.`, and record  special missing values as `.a` - `.z`.  These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.

@@

@@

<!-- Note: I think I'm basically there, but I'd like to revisit the first paragraph in this section and the last sentence -->

@@

The authors would like to thank Miles McBain, for his key contributions and discussions on the `naniar` package, in particular for helping implement `geom_miss_point`, and for his feedback on ideas, implementations, and names. We also thank Colin Fay for his contributions to the `naniar` package, in particular for his assistance with the `replace_with_na` functions. We also thank Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on missing data and package development, and for their assistance with creating elegant approaches that take advantage of the tidy syntax. We would also like to thank those who contributed pull requests and discussions on the `naniar` package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping `naniar` be more compliant with `ggplot2`. We would also like to thank Amelia McNamara for discussions on the paper.
